---
title: "Periodicity in daily test counts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Periodicity in daily test counts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(vdhcovid)
library(dplyr)
library(ggplot2)
```

The daily number of tests has a strong periodic signal.  
```{r dataprep}
apr15 <- as.Date('2020-04-15')
dailytotal <- 
  group_by(vadailytests, date) %>% 
  summarise(ntest=sum(ntest)) %>%
  filter(date >= apr15)
trendfit <- lm(ntest~date, dailytotal)
dailytotal$ntest_detrend <- dailytotal$ntest - predict(trendfit)
N <- nrow(dailytotal)
Nhalf <- floor(N/2 + 1)
ggplot(dailytotal, aes(x=date, y=ntest)) + geom_line(size=1.2, color='MidnightBlue') +
  theme_bw()
```

We can do a spectral analysis to work out what frequencies are represented.
```{r powerspec}
ps <- fft(dailytotal$ntest_detrend)[1:Nhalf]
f <- (seq_along(ps) - 1) / N
pltps <- data.frame(f=f, ps=Mod(ps))
ggplot(pltps, aes(x=f, y=ps)) + 
  geom_col(width=0.25*f[2], fill='MidnightBlue') + 
  ylab('Spectral density') +
  xlab('Frequency (1/day)') +
  theme_bw()
```

```{r perextract}
top3 <- order(pltps$ps, decreasing=TRUE)[1:3]
f3 <- f[top3]
per3 <- signif(1/f3, 2)
```
The top 3 modes have periods of `r paste(per3, collapse=', ')` days.  The first
of these confirms what we expected; there is a weekly periodicity that is probably
due to less testing and test processing on weekends.  The third is probably due
to weekends being two days long.  The middle one is about half a week; it's probably
related to the fact that the character of the periodicity is more like $\sin^2 t$
than $\sin t$.
