---
title: "Effective Reproduction Rate"
output: 
  html_document: default
  rmarkdown::html_vignette: default
  pdf_document: default
vignette: >
  %\VignetteIndexEntry{Effective Reproduction Rate}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  echo=FALSE,
  comment = "#>"
)
```

```{r setup, message=FALSE, warning=FALSE}
library(vdhcovid)
library(dplyr)
library(ggplot2)
library(ggthemes)
library(cowplot)
```

Estimates of effective reproduction rate $R_t$.  This is vdhcovid v`r packageVersion('vdhcovid')`.
Testing data is current through
`r max(vadailytests$date)`.  Case count data is current through `r max(vadailycases$date)`.

## Intermediate results

```{r adjustedcases}
## First, adjust cases counts for the number of tests done.  We only know the test 
## counts at the health district level, so we'll just use that adjustment for all
## localities in the district.
countycases <- select(vadailycases, date, fips, locality, HealthDistrict, cumcases=cases)
countycases <- group_by(countycases, fips) %>% mutate(cases = c(cumcases[1], diff(cumcases)))
districttests <- 
  mutate(vadailytests, fpos = npos / ntest) %>%
  select(date, HealthDistrict, ntest, fpos, npos)
countycases <- 
  left_join(countycases, districttests, by=c('date','HealthDistrict')) %>%
  filter(!is.na(ntest))
countycases$t <- as.numeric(countycases$date - as.Date('2020-01-01'))
adjustment <- 1000/countycases$ntest
countycases$adjusted_cases <- adjustment * countycases$cases
countycases$weight <- 1/adjustment
countycases$ladjcase <- log(pmax(countycases$adjusted_cases, 0.1))
```

```{r districtstats}
districtcases <- group_by(countycases, HealthDistrict, date) %>%
  summarise(cases=sum(cases), ntest=ntest[1], npos=npos[1], fpos=fpos[1])
districtcases$t <- as.numeric(districtcases$date - as.Date('2020-01-01'))
adjustment <- 1000 / districtcases$ntest
districtcases$adjusted_cases <- adjustment * districtcases$cases
districtcases$weight <- 1/adjustment
districtcases$ladjcase <- log(pmax(districtcases$adjusted_cases, 0.1))
```

### Smoothed adjusted cases

Plots of daily new cases, adjusted for the number of tests performed.  Each line is
a locality, and localities are grouped by health district.  Cases are
recorded at the locality level, while number of tests is recorded at the district
level, so the units here are cases in the locality per thousand cases in the district.
For districts that comprise a single locality, like Arlington, this is equivalent 
to the test positivity rate in units of permille.  For multi-locality districts, the
plotted figure will be less than the actual test positivity rate, since the number 
of cases is being divided by the number of tests done in the whole district, not
just the number of tests done in the locality.


```{r smoothing}
## Fit loess smoothing to log of adjusted cases.  We will use this to estimate 
## growth rates.
smoothingspan <- 0.5
smooth_county <- function(df, group) {
  fit <- loess(ladjcase~t, df, df$weight, span=smoothingspan)
  df$smoothlogcase <- predict(fit)
  df$smoothcase <- exp(df$smoothlogcase)
  df
}
countycases <- 
  group_by(countycases, fips) %>% 
  group_modify(smooth_county)

districtcases <-
  group_by(districtcases, HealthDistrict) %>%
  group_modify(smooth_county)

```

```{r plotadjcases, fig.height=9}
ggplot(countycases, aes(x=date, y=smoothcase, color=locality)) + geom_line(show.legend = FALSE) + facet_wrap(~HealthDistrict, scales='free_y', ncol=4) + ylab('Smoothed adjusted cases') + theme_bw()

```

### Raw growth rates

Growth rates (i.e., derivative of log cases), unadjusted for sample enrichment
effects.

```{r growthrates, fig.height=9}
county_growth <- function(df, group) {
  N <- nrow(df)
  i <- seq(1, N)
  ilo <- pmax(1, i-1)
  ihi <- pmin(N, i+1)
  df$growthrate <- (df$smoothlogcase[ihi]-df$smoothlogcase[ilo]) / (df$t[ihi]-df$t[ilo])
  df
}
countycases <-
  group_by(countycases, fips) %>%
  group_modify(county_growth) %>%
  mutate(pctgr = (exp(growthrate)-1)*100)

districtcases <-
  group_by(districtcases, HealthDistrict) %>%
  group_modify(county_growth)

ggplot(countycases, aes(x=date, y=pctgr, color=locality)) + geom_line(show.legend = FALSE) + facet_wrap(~HealthDistrict, scales='free_y', ncol=4) + ylab('Growth Rate (pct)') + theme_bw()

#ggplot(filter(districtcases, HealthDistrict=='Thomas Jefferson'), aes(x=date, y=pctgr, color=locality)) +
  

```

### Adjusted growth rates

These growth rates are adusted for enrichment.  We assume that overall prevalence
is low enough that we can use the approximation in our paper, for which the adjustment
depends only on the observed positive test fraction.  We use the positive test fraction
for each health district to adjust all of the localities within that district.  
In the early days the number of tests was small, and we would sometimes see 1/1 
positive tests for a 100% positive test fraction.  To deal with this, we cap the
positive test fraction at 50%.  At no time since the number of tests climbed to 
more than a hundred a day has it been higher than this anywhere. (Unlike the raw
growth rates, these have not been converted to percentages.)

```{r adjgrowth, fig.height=9}
countycases$adjgrowth <- countycases$growthrate / (1 - pmin(0.5, countycases$fpos))
districtcases$adjgrowth <- districtcases$growthrate / (1 - pmin(0.5, districtcases$fpos))

ggplot(districtcases, aes(x=date, y=adjgrowth)) + geom_line(size=1.2) +
  facet_wrap(~HealthDistrict, scales='free_y', ncol=4) + ylab('Growth Rate')+ theme_bw()

ggplot(countycases, aes(x=date, y=adjgrowth, color=locality)) + geom_line(show.legend = FALSE) + facet_wrap(~HealthDistrict, scales='free_y', ncol=4) + ylab('Growth Rate') + theme_bw()
```

## Effective reproduction number

We calculate the effective reproduction number from the adjusted growth rates, using
the equation $R_t = 1 + kD$, where $D$ is the serial interval.  To do this, we 
have to estimate a value for $D$.  Estimates of $D$ vary dramatically in the literature.
Here we use $D=6$, which is the value favored by our model and is within the range
of uncertainties reported in the literature.  Occasionally regions show extremely
high growth rates for short periods of time, typically early in the infection when
case counts are low.  We believe such measurements are artifacts; therefore, we 
cut off the display for $R_t$ values at 3.5.

### District aggregate Rt plots
```{r rtvals, fig.width=10}
D <- 6
countycases$rt <- pmax(0, 1 + countycases$adjgrowth * D)
districtcases$rt <- pmax(0, 1 + districtcases$adjgrowth * D)
alldistricts <- unique(districtcases$HealthDistrict)

### Make plots of locality Rt, grouped by district
### (can also be used for district plots)
plotrtrgn <- function(district, alldata, locality=TRUE)
{
  dat <- filter(alldata, HealthDistrict==district)
  if(locality) {
    pltbase <- ggplot(dat, aes(x=date, y=rt, color=locality))
  }
  else {
    pltbase <- ggplot(dat, aes(x=date, y=rt)) + geom_hline(yintercept=1, color='blue')
  }
  
  pltbase + geom_line(size=1) + 
    coord_cartesian(ylim=c(0,3.5)) +
    ggtitle(district) +
    theme_bw() 
}

ggplot(districtcases, aes(x=date, y=rt)) + geom_line(size=1) + 
  geom_hline(yintercept=1, color='blue') +
  coord_cartesian(ylim=c(0,3.5)) + 
  facet_wrap(~HealthDistrict) + 
  theme_bw(10)

```

Blow-ups of the individual districts:

```{r rtvalsdist}
distplots <- lapply(alldistricts, plotrtrgn, alldata=districtcases, locality=FALSE)
for(plt in distplots) {
  print(plt)
}
```

### Locality Rt plots

```{r rtvalslocal}
alldistricts <- unique(countycases$HealthDistrict)
localplots <- lapply(alldistricts, plotrtrgn, alldata=countycases)
for(plt in localplots) {
  print(plt)
}
```
